---
title: "Maternal Mortality"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r, message = FALSE, warning = FALSE}
#install.packages("tidyverse")
library(tidyverse)
library(ggridges)
library(readxl)
library(patchwork)
#install.packages("plotly")
library(plotly)
#install.packages("ggtext")
library(ggtext)
library(tidyr)
library(readr)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)
theme_set(theme_minimal() + theme(legend.position = "bottom"))
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

```


## Maternal Mortality factors as a predictor of Death

Maternal Mortality has been a cause of death for mothers due to a variety of factors that seem preventable or modifiable. To determine the underlying or root of these causes.

```{r}

maternalmorb<- read_csv("data/Pregnancy-Associated_Mortality.csv")

mdeathsbyrace = maternalmorb %>%
  ggplot(aes(x = `Race/ethnicity`, y = Deaths, fill = `Race/ethnicity`)) + 
  geom_boxplot() + 
  labs(title = "Maternal Deaths by Race")

mdeathsbyrace

```

```{r}
maternalmorb<- read_csv("data/Pregnancy-Associated_Mortality.csv")

mdeathsbyborough = maternalmorb %>%
  filter(Borough != "All") %>%
  ggplot(aes(x = Borough, y = Deaths, fill = Borough)) + 
  geom_boxplot() + 
  labs(title = "Maternal Deaths by Borough")

mdeathsbyborough


```

```{r}
maternalmorb<- read_csv("data/Pregnancy-Associated_Mortality.csv")

mdeathsunderlyingcause = maternalmorb %>%
  filter(Related == "Pregnancy-related") %>%
  ggplot(aes(x = Underlying_cause, y = Deaths, fill= Underlying_cause)) + 
  geom_boxplot() + 
  labs(title = "Maternal Deaths by Underlying Cause") + 
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust=1))

mdeathsunderlyingcause

```

## Discussion 

There is a clear discrepancy in maternal mortality based on race. Black women are more likely to die due a pregnancy related cause than any other race or ethnic group. 
Other or Biracial women have the lowest pregnancy related mortality within the 2016 - 2019 time period. 

## Limitations

These data were restricted to to availability and completeness of the data resulting in less in depth analysis for other boroughs by race and further examination of the underlying causes. 
