---
title: "Maternal Mortality"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r include = FALSE}
library(tidyverse)
library(dplyr)
library(patchwork)
library(leaflet)
library(lubridate)
library(rvest)
library(httr)
library(flexdashboard)
library(plotly)
library(viridis)
library(stringr)
knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)
theme_set(theme_minimal() + theme(legend.position = "bottom"))
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```


Maternal Mortality Data in New York City [(Dataset)](https://data.cccnewyork.org/data/download#0,4,8,13/0,1280) 

Maternal Mortality Data in New York City and its boroughs was obtained from an organization called Citizens' Committee for Children of New York which organized tabular data on maternal mortality between 2016 - 2019, for all neighborhoods and boroughs in New York City. To clean the data, we had to remove columns that were not relevant to the analysis or contained explanations for the data formats. We also had to rename columns to their appropriate titles as well as converting column types to numeric forms as needed. 

The columns of interest for our analysis are: 

`borough`: The borough or neighborhood within New York City 

`race/ethnicity`: The race/ethnicity of the individual.

`underlying cause`: Pregnancy associated underlying causes of mortality.

`year`: The year the data was reported 


```{r, message = FALSE, warning = FALSE}

maternalmorb<- read_csv("data/Pregnancy-Associated_Mortality.csv")

mdeathsbyrace = maternalmorb %>%
  ggplot(aes(x = `Race/ethnicity`, y = Deaths, fill = `Race/ethnicity`)) + 
  geom_boxplot() + 
  labs(title = "Maternal Deaths by Race")

mdeathsbyrace
```

```{r}
maternalmorb<- read_csv("data/Pregnancy-Associated_Mortality.csv")

mdeathsbyborough = maternalmorb %>%
  filter(Borough != "All") %>%
  ggplot(aes(x = Borough, y = Deaths, fill = Borough)) + 
  geom_boxplot() + 
  labs(title = "Maternal Deaths by Borough")

mdeathsbyborough
```

```{r}
maternalmorb<- read_csv("data/Pregnancy-Associated_Mortality.csv")

mdeathsunderlyingcause = maternalmorb %>%
  filter(Related == "Pregnancy-related") %>%
  ggplot(aes(x = Underlying_cause, y = Deaths, fill= Underlying_cause)) + 
  geom_boxplot() + 
  labs(title = "Maternal Deaths by Underlying Cause") + 
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust=1))

mdeathsunderlyingcause
```
